---
title: "Keeping tabs on taxing agencies"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

# Introduction

One of the unique features of PTAXSIM is the database that accompanies the package - this DB contains the only publicly accessible, machine-readable data collected from the various Cook County property tax agencies going back to 2006. While the primary use of PTAXSIM is its functionality to calculate tax bills, its database holds a plethora of data that can be used to investigate Cook County's property tax system, including the behavior of the over 900 taxing agencies and 400 TIFs that collect revenue through property taxes.

In this vignette, we'll demonstrate how to query data from the PTAXSIM SQLite database that can help us analyze taxing agencies' property tax revenue over time, as well as revenue collected by TIFs.

Additionally, we'll show how to account for the 2024 changes to the Cook County Clerk's agency fund report data structure which will be necessary for users to pay attention to when conducting a time series analysis of certain taxing agencies.

# Chicago's levy over time

Using data from the PTAXSIM database, let's look at the levy history for the City of Chicago from 2006 to 2024.

First, load some useful libraries and instantiate a PTAXSIM DBI connection with the default name (`ptaxsim_db_conn`) expected by PTAXSIM functions.

```{r}
library(data.table)
library(dplyr)
library(DT)
library(here)
library(ggplot2)
#library(ptaxsim)
library(tidyr)
devtools::load_all()

ptaxsim_db_conn <- DBI::dbConnect(RSQLite::SQLite(), here("./ptaxsim.2.db"))
```

```{r, echo=FALSE}
# This is needed to build the vignette using GitHub Actions
#ptaxsim_db_conn <- DBI::dbConnect(
 # RSQLite::SQLite(),
  #Sys.getenv("PTAXSIM_DB_PATH")
#)
```

First, we'll query the database table `agency_info` to determine the City of Chicago's unique `agency_num`.

```{r}
chi_agency_nums <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT agency_num, agency_name
  FROM agency_info
  WHERE agency_name LIKE '%CITY OF CHICAGO%'
  "
)

head(chi_agency_nums)
```
For purpose of our analysis, we'll just focus on `CITY OF CHICAGO` and `CITY OF CHICAGO LIBRARY FUND`, with agency fund numbers `030210000` and `030210002` respectively. While the `CITY OF CHICAGO LIBRARY FUND` has been treated as its own taxing agency in the Clerk's data, the City itself reports the amount levied for the purpose of funding the Chicago Public Library as a portion of its entire property tax levy.

First, we'll query all fields from the `agencies` table.

```{r}
chi_agencies <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT DISTINCT *
  FROM agency
  WHERE agency_num = '030210000'
  OR agency_num = '030210001'
  "
)
```

## Accounting for 2024 changes to agency fund reporting

When examining `chi_agencies`, we see that in 2024 the `CITY OF CHICAGO LIBRARY FUND` has a \$0 levy and therefore a tax rate of 0. This is because, starting in 2024, the Cook County Clerk began reporting the requested levy for the City of Chicago libraries as a fund under the `CITY OF CHICAGO` taxing agency. As mentioned above, this is in fact aligned with how the City of Chicago reports its property tax levy in its own budget [documentation](https://chicityclerk.s3.us-west-2.amazonaws.com/s3fs-public/O2023-0005291_Tax_Levy.pdf).

This reporting change occurred for many municipal taxing agencies, where the Clerk had previously reported municipal fund levies as separate taxing agencies. This updated data structure begins in 2024 while years prior to 2024 remain the same, meaning users will need to account for this discrepancy if ever analyzing these taxing agencies data over time.

To make this process easier, we have added new fields to the `agency_info` table in the PTAXSIM database which identify the agencies that have been folded into their parent agencies. `agency_num_24` and `agency_name_24` contain the agency info that the old "sub-agency" has been merged into. Note that the user can still see details about these former agencies, now funds, by querying the `agency_fund` table.

The table below shows all of the taxing agencies that have been been folded into their parent agencies beginning with tax year 2024. In all, we have 78 former taxing agencies that are no longer reported as individual agencies beginning in 2024.

```{r}
# Query agency_info table for all agencies with the 2024 update
agency_cw_24 <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT *
  FROM agency_info
  WHERE agency_change_24 = 1
  "
) %>%
  select(agency_num, agency_name, agency_num_24, agency_name_24)

datatable(agency_cw_24)

```
We can now use `agency_cw_24` as a crosswalk to convert these sub-agencies to their parent agency numbers. 

Knowing this caveat about 2024 agency data, we'll query the both `CITY OF CHICAGO` and `CITY OF CHICAGO LIBRARY FUND` for all years, and then update the `agency_num` field for `CITY OF CHICAGO LIBRARY FUND` using `agency_cw_24`. This will ensure consistent reporting across all years.

```{r}
chi_agencies <- chi_agencies %>%
  # Join the agency crosswalk to get the parent agency number for pre-2024 years
  left_join(agency_cw_24, "agency_num") %>%
  # for the agencies that did have an agency number change in 2024, replace the
  # old agency_num with the new one
  mutate(
    agency_num =
      ifelse(!is.na(agency_num_24),
             agency_num_24,
             agency_num)
  )
```

## Tracking City of Chicago's levy growth with inflation

Now that we have the correct agency numbers for all years, we can select and aggregate the fields of interest for `CITY OF CHICAGO`. 

Many of the fields in the `agencies` table, which come from the Clerk's agency rate report, relate to PTELL calculations. Because the Chicago is a home rule municipality, it is not subject to PTELL (Property Tax Extension Law Limit) as imposed by the State of Illinois, so these fields will not contain relevant info. However, the City of Chicago imposes [its own limits](https://codelibrary.amlegal.com/codes/chicago/latest/chicago_il/0-0-0-2608573) which mirror PTELL's limits which prohibit a taxing agency from increasing its levy more than the rate of inflation or 5%, whichever is less. 

```{r}
chi_levy <- 
  chi_agencies %>%
  group_by(year, agency_num) %>%
  summarize(
    total_final_levy = sum(total_final_levy),
    total_final_rate = sum(total_final_rate),
    total_ext = sum(total_ext),
    cty_cook_eav = first(cty_cook_eav),
    prior_eav = first(prior_eav),
    curr_new_prop = first(curr_new_prop)
  )
```
PTELL, as well as Chicago's property tax limitation ordinance, contain loopholes that allow taxing agencies to [increase their levies beyond the rate of inflation](https://civicfed.org/press/new-report-finds-illinois-property-tax-cap-law-not-working-intended). Let's see how Chicago's levy has fared compared to the inflation rate since 2006.

To do this we can pull CPI data from the PTAXSIM database - we include the CPI tables published annually by the Illinois Department of Revenue (IDOR) for the purpose of setting the PTELL limit.

```{r}
cpi <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT 
  levy_year as year,
  cpi,
  ptell_cook
  FROM cpi
  "
) %>%
  filter(year >= 2006)
```

First, let's convert each year's levy to be in terms of 2006 dollars using the consumer price index (CPI). This will help us understand the degree to which the levy has been adjusted on account of inflation vs. increasing revenue.

```{r}
cpi <- cpi %>%
  mutate(multiplier = cpi/cpi[year == 2006])

chi_levy_adj <- chi_levy %>%
  left_join(cpi, "year") %>%
  mutate(
    total_final_levy_adj = total_final_levy / multiplier
  ) 

```

Plotting Chicago's total property tax levy from 2006 to 2024 in both 2006 dollars and nominal dollars, we can see that the levy remained steady from 2006 to 2014. The levy actually decreased slightly in real dollars given the levy was not increased to account for inflation during this time. Then in 2015, City begins increasing its levy significantly due to State legislation that dictates [funding requirements for Chicago's Fire and Police pension funds](https://www.civicfed.org/civic-federation/blog/chicago-police-and-fire-pension-funding-changes-become-law).

Rising inflation beginning in the wake of the Covid-19 pandemic is evident with the levy in "real" dollars decreasing slightly from 2020 to 2024 even as the levy in nominal dollars increased in that time period.

<details>

<summary><strong>Click here</strong> to show plot code</summary>
```{r}
chi_levy_plot_1 <- chi_levy_adj %>%
  ggplot() +
  geom_line(aes(x = year, y = total_final_levy, color = "Levy in nominal dollars"),
            linewidth = .8) +
  geom_point(aes(x = year, y = total_final_levy, color = "Levy in nominal dollars")) +
  geom_line(aes(x = year, y = total_final_levy_adj, color = "Levy in 2006 dollars"),
            linewidth = .8, linetype = "dashed") +
  geom_point(aes(x = year, y = total_final_levy_adj, color = "Levy in 2006 dollars")) +
  scale_color_manual(
    name = NULL,
    values = c("Levy in nominal dollars" = "black",
               "Levy in 2006 dollars"    = "blue")
  ) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(
    labels = scales::label_dollar(scale = 1e-9, suffix = "B"),
    limits = c(0, 2000000000)
  ) +
  labs(
    x = NULL,
    y = "Total Final Levy (Billions)"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 13),
    axis.title.x = element_text(margin = margin(t = 6)),
    axis.title.y = element_text(margin = margin(r = 6)),
    axis.text = element_text(size = 11),
    strip.text = element_text(size = 16),
    strip.background = element_rect(fill = "#c9c9c9"),
    legend.title = element_text(size = 14),
    legend.key.size = unit(24, "points"),
    legend.text = element_text(size = 12),
    legend.position = "bottom"
  )

```

</details>

<br>

```{r, echo=FALSE, out.width="100%"}
chi_levy_plot_1
```
Next, we'll take a look at how the percent change in the levy compares to the percent change in CPI - or the rate of inflation. We add some data manipulation to calculate the percent change in both the levy and CPI by each year since 2006.

```{r}
chi_levy_adj <- chi_levy_adj %>%
  ungroup() %>%
  mutate(
    pct_change_final_levy     = (total_ext - total_ext[year == 2006]) / total_ext[year == 2006] * 100,
    pct_change_ptell          = (1 * c(1, cumprod(1 + tail(ptell_cook, -1))) - 1) * 100,
  )
```

Plotting the percent change of the City's levy since 2006, we see it has outgrown the cumulative effects of the City's property tax limitation 

<details>

<summary><strong>Click here</strong> to show plot code</summary>
```{r, echo=FALSE, out.width="100%"}
chi_levy_plot_2 <- chi_levy_adj %>%
  pivot_longer(
    cols = c(pct_change_final_levy, pct_change_ptell),
    names_to = "series",
    values_to = "pct_change"
  ) %>%
  mutate(series = recode(series,
    "pct_change_final_levy"     = "Levy % change (nominal $)",
    "pct_change_ptell"            = "PTELL % change"
  )) %>%
  ggplot() +
  geom_line(aes(x = year, y = pct_change, color = series), linewidth = .8) +
  geom_point(aes(x = year, y = pct_change, color = series)) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(labels = scales::label_percent(scale = 1, suffix = "%")) +
  labs(
    x = NULL,
    y = "Percent Change Since Base Year",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 13),
    axis.title.y = element_text(margin = margin(r = 6)),
    axis.text = element_text(size = 11),
    strip.text = element_text(size = 16),
    strip.background = element_rect(fill = "#c9c9c9"),
    legend.title = element_text(size = 14),
    legend.key.size = unit(24, "points"),
    legend.text = element_text(size = 12),
    legend.position = "bottom"
  )
```
</details>

<br>

```{r, echo=FALSE, out.width="100%"}
chi_levy_plot_2
```


```{r}
chi_agency_fund <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT *
  FROM agency_fund
  WHERE agency_num = '030210000'
  OR agency_num = '030210001'
  "
)

chi_agency_fund_info <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT *
  FROM agency_fund_info
  WHERE agency_num = '030210000'
  OR agency_num = '030210001'
  "
)

chi_agency_fund <- chi_agency_fund %>%
  left_join(chi_agency_fund_info) %>%
  filter(final_levy > 0,
         fund_type_num != "319") %>%
  mutate(
    fund_catg =
      case_when(grepl("A & B", fund_name) ~ "Pension",
                grepl("NOTE REDEMPTION & INTEREST FUND", fund_name) ~ 
                  "Bond Payments",
                grepl("BONDS & INTEREST", fund_name) ~ "Bond Payments",
                grepl("LIBRARY NOTE REDEMPTION", fund_name) ~ "Library Fund",
                grepl("BONDS & INTEREST", fund_type_name) ~ "Bond Payments",
                grepl("LIBRARY NOTE REDEMPTION", fund_type_name) ~ "Library Fund",
                TRUE ~ fund_type_name)
  ) %>%
  group_by(year, fund_catg) %>%
  summarise(final_levy = sum(final_levy),
             final_rate = sum(final_rate))

```

<details>

<summary><strong>Click here</strong> to show plot code</summary>
```{r}
chi_levy_plot_3 <- chi_agency_fund %>%
  ggplot() +
  geom_line(aes(x = year, y = final_levy, color = fund_catg), 
            linewidth = .8) +
  geom_point(aes(x = year, y = final_levy, color = fund_catg)) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(
    labels = scales::label_dollar(scale = 1e-9, suffix = "B"),
    limits = c(0, NA)
  ) +
  labs(
    x = NULL,
    y = "Final Levy (Billions)",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 13),
    axis.title.y = element_text(margin = margin(r = 6)),
    axis.text = element_text(size = 11),
    strip.text = element_text(size = 16),
    strip.background = element_rect(fill = "#c9c9c9"),
    legend.title = element_text(size = 14),
    legend.key.size = unit(24, "points"),
    legend.text = element_text(size = 12),
    legend.position = "bottom"
  )
```
</details>

<br>

```{r, echo=FALSE, out.width="100%"}
chi_levy_plot_3
```

```{r}
fund_type_table <- chi_agency_fund %>%
  select(-final_rate) %>%
  filter(year > 2019) %>%
  pivot_wider(names_from = year, values_from = final_levy)

fund_type_table %>%
  arrange(desc(`2024`)) %>%
  bind_rows(
    summarise(., fund_catg = "Total Levy", across(c("2020", "2021", "2022", "2023", "2024"), sum))
  ) %>%
    rename(`Fund Type` = fund_catg) %>%
  datatable(
    caption = "Chicago Property Tax Levies by Fund Type",
    rownames = FALSE,
    options = list(
      pageLength = nrow(fund_type_table) + 1,
      dom = "t",
      columnDefs = list(
        list(className = "dt-left", targets = 0),
        list(className = "dt-right", targets = 1:5)
      )
    )
  ) %>%
  formatCurrency(
    columns = c("2020", "2021", "2022", "2023", "2024"),
    currency = "$",
    digits = 0
  )
```
```{r}
fund_type_table %>%
  mutate(across(c("2020", "2021", "2022", "2023", "2024"), 
                ~ round(. / sum(.) * 100))) %>%
  arrange(desc(`2024`)) %>%
  datatable(
    caption = "Chicago Property Tax Levies by Fund Type — Share by Year",
    rownames = FALSE,
    options = list(
      pageLength = nrow(fund_type_table),
      dom = "t",
      columnDefs = list(
        list(className = "dt-left", targets = 0),
        list(className = "dt-right", targets = 1:5)
      )
    )
  ) %>%
  formatRound(
    columns = c("2020", "2021", "2022", "2023", "2024"),
    digits = 1
  ) %>%
  formatString(
    columns = c("2020", "2021", "2022", "2023", "2024"),
    suffix = "%"
  )
```
```{r}
chi_levy_plot_4 <- chi_levy %>%
  ggplot() +
  geom_line(aes(x = year, y = total_final_rate), linewidth = .8) +
  geom_point(aes(x = year, y = total_final_rate)) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(labels = scales::label_percent(scale = 1, suffix = "%"),
                     limits = c(0, 2)) +
  labs(
    x = NULL,
    y = "Tax Rate"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 13),
    axis.title.y = element_text(margin = margin(r = 6)),
    axis.text = element_text(size = 11),
    strip.text = element_text(size = 16),
    strip.background = element_rect(fill = "#c9c9c9"),
    legend.title = element_text(size = 14),
    legend.key.size = unit(24, "points"),
    legend.text = element_text(size = 12),
    legend.position = "bottom"
  )
```

```{r}
chi_levy_plot_4
```

