---
title: "Keeping tabs on taxing agencies"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

# Introduction

One of the unique features of PTAXSIM is the database that accompanies the package - this DB contains the only publicly accessible, machine-readable data collected from the various Cook County property tax agencies going back to 2006. While the primary use of PTAXSIM is its functionality to calculate tax bills, its database holds a plethora of data that can be used to investigate Cook County's property tax system, including the behavior of the over 900 taxing agencies and 400 TIFs that collect revenue through property taxes.

In this vignette, we'll demonstrate how to query data from the PTAXSIM SQLite database that can help us analyze taxing agencies' property tax revenue over time, as well as revenue collected by TIFs.

Additionally, we'll show how to account for the 2024 changes to the Cook County Clerk's agency fund report data structure which will be necessary for users to pay attention to when conducting a time series analysis of certain taxing agencies.

# Chicago taxing agencies

Using data from the PTAXSIM database, let's look at the levy history for the primary taxing agencies that appear on a Chicago property's tax bill: the `CITY OF CHICAGO`, `BOARD OF EDUCATION` (a.k.a. Chicago Public Schools), and `CHICAGO PARK DISTRICT`.

First, load some useful libraries and instantiate a PTAXSIM DBI connection with the default name (`ptaxsim_db_conn`) expected by PTAXSIM functions.

```{r}
library(data.table)
library(dplyr)
library(DT)
library(here)
library(ggplot2)
#library(ptaxsim)
library(tidyr)
devtools::load_all()

ptaxsim_db_conn <- DBI::dbConnect(RSQLite::SQLite(), here("./ptaxsim.2024.0.0-alpha.2.db"))
```

```{r, echo=FALSE}
# This is needed to build the vignette using GitHub Actions
#ptaxsim_db_conn <- DBI::dbConnect(
 # RSQLite::SQLite(),
  #Sys.getenv("PTAXSIM_DB_PATH")
#)
```

We can query the database table `agency_info` to search for `agency_num` key assigned the the taxing agencies of interest. This is what shows up when we search of agencies with the `CHICAGO` in the `agenc_name`.

```{r}
chi_agency_nums <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT agency_num, agency_name
  FROM agency_info
  WHERE agency_name LIKE '%CHICAGO%'
  "
)

head(chi_agency_nums)
```

Note that there are several agencies that pop up when we search of agencies that contain `CHICAGO` in the name. Let's take a moment to examine the agencies `CITY OF CHICAGO` and `CITY OF CHICAGO LIBRARY FUND`, with agency fund numbers `030210000` and `030210002` respectively. While the `CITY OF CHICAGO LIBRARY FUND` has been treated as its own taxing agency in the Clerk's data, the City itself reports the amount levied for the purpose of funding the Chicago Public Library as a portion of its entire property tax levy.

We're also interested in the `BOARD OF EDUCATION` and `CHICAGO PARK DISTRICT` agencies, which have agency numbers `050200000` and `044060000` respectively. We'll query all fields from the `agency` table and filter by their `agency_num`. 

```{r}
chi_agencies <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT DISTINCT *
  FROM agency
  WHERE agency_num = '030210000'
  OR agency_num = '030210001'
  OR agency_num = '050200000'
  OR agency_num = '044060000'
  "
)
```

## Accounting for 2024 changes to agency fund reporting

When examining `chi_agencies`, we see that in 2024 the `CITY OF CHICAGO LIBRARY FUND` has a \$0 levy and therefore a tax rate of 0. This is because, starting in 2024, the Cook County Clerk began reporting the requested levy for the City of Chicago libraries as a fund under the `CITY OF CHICAGO` taxing agency. As mentioned above, this is in fact aligned with how the City of Chicago reports its property tax levy in its own budget [documentation](https://chicityclerk.s3.us-west-2.amazonaws.com/s3fs-public/O2023-0005291_Tax_Levy.pdf).

This reporting change occurred for many municipal taxing agencies, where the Clerk had previously reported municipal fund levies as separate taxing agencies. This updated data structure begins in 2024 while years prior to 2024 remain the same, meaning users will need to account for this discrepancy if ever analyzing these taxing agencies data over time.

To make this process easier, we have added new fields to the `agency_info` table in the PTAXSIM database which identify the agencies that have been folded into their parent agencies. `agency_num_24` and `agency_name_24` contain the agency info that the old "sub-agency" has been merged into. Note that the user can still see details about these former agencies, now funds, by querying the `agency_fund` table.

The table below shows all of the taxing agencies that have been been folded into their parent agencies beginning with tax year 2024. In all, we have 78 former taxing agencies that are no longer reported as individual agencies beginning in 2024.

```{r}
# Query agency_info table for all agencies with the 2024 update
agency_cw_24 <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT *
  FROM agency_info
  WHERE agency_change_24 = 1
  "
) %>%
  select(agency_num, agency_name, agency_num_24, agency_name_24)

datatable(agency_cw_24)
```
We can now use `agency_cw_24` as a crosswalk to convert these sub-agencies to their parent agency numbers. 

Knowing this caveat about 2024 agency data, we'll query the both `CITY OF CHICAGO` and `CITY OF CHICAGO LIBRARY FUND` for all years, and then update the `agency_num` field for `CITY OF CHICAGO LIBRARY FUND` using `agency_cw_24`. This will ensure consistent reporting across all years.

```{r}
chi_agencies <- chi_agencies %>%
  # Join the agency crosswalk to get the parent agency number for pre-2024 years
  left_join(agency_cw_24, "agency_num") %>%
  # for the agencies that did have an agency number change in 2024, replace the
  # old agency_num with the new one
  mutate(
    agency_num =
      ifelse(!is.na(agency_num_24),
             agency_num_24,
             agency_num)
  )
```

## Tracking Chicago levies

Now that we have the correct agency numbers for all years, we can select and aggregate the fields of interest for the Chicago taxing agencies we want to look at.

Let's create an object called `chi_levy` that contains final total levy, taxable aggregate EAV (or the tax base), and the amount of EAV designated as "New Property". New Property is defined as EAV from expiring TIFs, new constriction, or expiring incentives.

```{r}
chi_levy <- 
  chi_agencies %>%
  group_by(year, agency_num) %>%
  summarize(
    total_final_levy = sum(total_final_levy),
    total_final_rate = sum(total_final_rate),
    total_ext = sum(total_ext),
    cty_cook_eav = first(cty_cook_eav),
    prior_eav = first(prior_eav),
    curr_new_prop = first(curr_new_prop)
  )

chi_levy_indx <- chi_levy %>%
  group_by(agency_num) %>%
  mutate(levy_2006 = total_final_levy[year == 2006][1],
         levy_pct_inc = (total_final_levy - levy_2006) / levy_2006 * 100) %>%
  ungroup() %>%
  select(agency_num, year, levy_pct_inc) %>%
  pivot_longer(
    cols = levy_pct_inc,
    names_to = "series",
    values_to = "pct_inc"
  )
```

Many of the fields in the `agencies` table, which come from the Clerk's agency rate report, relate to the Clerk's PTELL (Property Tax Extension Law Limit) calculations. PTELL ensures certain taxing agencies do not increase their levies beyond the rate of inflation (with some exceptions). CPS and Chicago Park District are subject to PTELL rules, but the City of Chicago is not by virtue of being a non-home rule agency. However, the City of Chicago imposes [its own limits](https://codelibrary.amlegal.com/codes/chicago/latest/chicago_il/0-0-0-2608573) which mirror PTELL's limits which prohibit a taxing agency from increasing its levy more than the rate of inflation or 5%, whichever is less. 

PTELL, as well as Chicago's property tax limitation ordinance, contain loopholes that allow taxing agencies to [increase their levies beyond the rate of inflation](https://civicfed.org/press/new-report-finds-illinois-property-tax-cap-law-not-working-intended). Let's see how Chicago's levy has fared compared to inflation  since 2006.

To do this we can pull CPI data from the PTAXSIM database - we include the CPI tables published annually by the Illinois Department of Revenue (IDOR) for the purpose of setting the PTELL limit. We will calculate the CPI percent change from 2006, and then combine that data with the levy percent change data from `chi_levy_indx`. 

```{r}
cpi <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT *
  FROM cpi
  "
) %>%
  mutate(
    pct_inc = (cpi / cpi[year == 2006] - 1) * 100,
    agency_num = "100",
    series = "cpi"
  ) %>%
  filter(year >= 2006) %>%
  select(year, agency_num, series, pct_inc)
```


<details>

<summary><strong>Click here</strong> to show plot code</summary>
```{r}
highlight_key <- tibble::tibble(
  agency_num = c("100", "050200000", "030210000", "044060000"),
  label      = c("CPI-U", "CPKD",      "City",        "CPS")
)

plot_1_df <- rbind(cpi, chi_levy_indx) %>%
  left_join(highlight_key, by = "agency_num")
  
df_lab <- plot_1_df %>%
  group_by(agency_num, label) %>%
  filter(year == max(year, na.rm = TRUE)) %>%
  slice_tail(n = 1) %>%       
  ungroup() %>%
  mutate(label_col = ifelse(label == "CPI-U", "#d62728", "black"))  # red for CPI-U

chi_levy_plot_1 <- 
  ggplot() +
  geom_line(
    data = plot_1_df,
    aes(x = year, y = pct_inc/100, group = agency_num, color = label),
    linewidth = .5
  ) +
  # right-side labels in black (no color mapping)
  geom_text(
    data = df_lab,
    aes(x = year, y = pct_inc/100, label = label),
    color = df_lab$label_col,
    hjust = -0.1,
    size = 3.5
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(
    breaks = seq(min(plot_1_df$year, na.rm = TRUE), max(plot_1_df$year, na.rm = TRUE), by = 2),
    expand = expansion(mult = c(0.01, 0.08))  # add right padding so labels fit
  ) +
  scale_color_manual(
    values = c(
      "CPKD"   = "black",
      "City"   = "black",
      "CPS"    = "black",
      "CPI-U"  = "#d62728"
    ),
    breaks = c("CPKD", "City", "CPS"),
    guide = "none"  # hide legend since lines are labeled
  ) +
  labs(
    x = "Year",
    y = "Percent change from 2006",
  ) +
  theme_minimal(base_size = 12)
```
</details>

<br>

Next, we will plot the percent change in levy from 2006 for the City of Chicago, Chicago Public Schools, and Chicago Park District, and compare that to the percent change in CPI-U over the same period.

We can see that the City of Chicago's levy has increased the most, having more than doubled from 2006 to 2024. Both Chicago and CPS's levies have increased far beyond the rate of inflation, while the Chicago Park District's levy has kept pace with it.

```{r, echo=FALSE, out.width="100%"}
chi_levy_plot_1
```

Even though the City's levy has increased the most in percentage terms since 2006, it does not have the largest levy. The taxing agencies that rely most on property taxes are school districts and this is true for Chicago as well. The plot below shows the final tax extensions (meaning the final amount to be received by taxing agencies after the levies are validated by PTELL), for each of the three agencies. CPS is slightly more than double the City's extension in 2024 at about $4 Billion. Chicago Park District's extension is substantially less at less than 1 Billion. 

<summary><strong>Click here</strong> to show plot code</summary>
```{r}
plot_2_df <- chi_levy %>%
  left_join(highlight_key)

df_lab <- plot_2_df %>%
  group_by(agency_num, label) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup()

df_2024 <- plot_2_df %>%
  filter(year == 2024) %>%
  group_by(agency_num, label) %>%
  slice_tail(n = 1) %>%  # in case there are duplicates in 2024
  ungroup() %>%
  mutate(label_2024 = paste0(
    label, ": ", scales::label_dollar(scale = 1e-9, suffix = "B", accuracy = 0.1)(total_ext)
  ))

chi_levy_plot_2 <- ggplot(plot_2_df, aes(year, total_ext, group = agency_num)) +
  geom_line(linewidth = 0.5, color = "black") +
  # 2024 value labels like "CPS: $1.1B"
  geom_text(
    data = df_2024,
    aes(label = label_2024),
    color = "black",
    hjust = .6,
    vjust = -.5
  ) +
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-9, suffix = "B")) +
  scale_x_continuous(
    breaks = seq(min(plot_2_df$year, na.rm = TRUE), max(plot_2_df$year, na.rm = TRUE), by = 2),
    expand = expansion(mult = c(0.01, 0.08))
  ) +
  labs(x = "Year", y = "Final Agency Extension") +
  theme_minimal(base_size = 12)
```
</details>

<br>

```{r, echo=FALSE, out.width="100%"}
chi_levy_plot_2
```
The PTAXSIM database also contains information related to taxing agency's property tax funds so we can understand exactly what they intend to spend their property tax revenue on. This information can be found in the `agency_fund` and `agency_fund_info` tables. This data is not utilized by any of PTAXSIM's functions, but it is available to be queried an analyzed. 

We will query the fund information for the `CITY OF CHICAGO` and the `CITY OF CHICAGO LIBRARY FUND`. 

To simplify the `agency_fund` data, we will add broader categories to define funds. For example, there is not a consistent identifier for which funds are contributions to the City's various pension funds. In this case, we'll tag any fund with "A & B" (Annuities and Benefits) in the name as a pension fund. We'll label funds related to bond and interest payments, as well as note redemption, as "Bond Payments".

```{r}
chi_agency_fund <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT *
  FROM agency_fund
  WHERE agency_num = '030210000'
  OR agency_num = '030210001'
  "
)

chi_agency_fund_info <- DBI::dbGetQuery(
  ptaxsim_db_conn,
  "
  SELECT *
  FROM agency_fund_info
  WHERE agency_num = '030210000'
  OR agency_num = '030210001'
  "
)

chi_agency_fund <- chi_agency_fund %>%
  left_join(chi_agency_fund_info) %>%
  filter(final_levy > 0,
         fund_type_num != "319") %>%
  mutate(
    fund_catg =
      case_when(grepl("A & B", fund_name) ~ "Pension",
                grepl("NOTE REDEMPTION & INTEREST FUND", fund_name) ~ 
                  "Bond Payments",
                grepl("BONDS & INTEREST", fund_name) ~ "Bond Payments",
                grepl("LIBRARY NOTE REDEMPTION", fund_name) ~ "Library Fund",
                grepl("BONDS & INTEREST", fund_type_name) ~ "Bond Payments",
                grepl("LIBRARY NOTE REDEMPTION", fund_type_name) ~ "Library Fund",
                TRUE ~ fund_type_name)
  ) %>%
  group_by(year, fund_catg) %>%
  summarise(final_levy = sum(final_levy),
             final_rate = sum(final_rate))

```

The plot below 

<details>

<summary><strong>Click here</strong> to show plot code</summary>
```{r}
chi_levy_plot_3 <- chi_agency_fund %>%
  ggplot() +
  geom_line(aes(x = year, y = final_levy, color = fund_catg), 
            linewidth = .5) +
  geom_point(aes(x = year, y = final_levy, color = fund_catg)) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(
    labels = scales::label_dollar(scale = 1e-9, suffix = "B"),
    limits = c(0, NA)
  ) +
  labs(
    x = NULL,
    y = "Final Levy (Billions)",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 13),
    axis.title.y = element_text(margin = margin(r = 6)),
    axis.text = element_text(size = 11),
    strip.text = element_text(size = 16),
    strip.background = element_rect(fill = "#c9c9c9"),
    legend.title = element_text(size = 14),
    legend.key.size = unit(24, "points"),
    legend.text = element_text(size = 12),
    legend.position = "bottom"
  )
```
</details>

<br>

The plot below illustrates how the City's levy increase in 2016 was by soley driven by a massive increase in the City's pension contributions. Pension contributions grew again in 2021, but in tandem with a decrease in bond financing.

```{r, echo=FALSE, out.width="100%"}
chi_levy_plot_3
```
With so much of the City's property tax revenue going to funding pensions, it's difficult to see if or how there was any change to the amount going to the Library Fund. The below tables show the final levy for each fund type in 2020-2024, as well as the share of the total levy for each fund type by year. 

```{r, echo = FALSE}
fund_type_table <- chi_agency_fund %>%
  select(-final_rate) %>%
  filter(year > 2019) %>%
  pivot_wider(names_from = year, values_from = final_levy)

fund_type_table %>%
  arrange(desc(`2024`)) %>%
  bind_rows(
    summarise(., fund_catg = "Total Levy", across(c("2020", "2021", "2022", "2023", "2024"), sum))
  ) %>%
    rename(`Fund Type` = fund_catg) %>%
  datatable(
    caption = "Chicago Property Tax Levies by Fund Type",
    rownames = FALSE,
    options = list(
      pageLength = nrow(fund_type_table) + 1,
      dom = "t",
      columnDefs = list(
        list(className = "dt-left", targets = 0),
        list(className = "dt-right", targets = 1:5)
      )
    )
  ) %>%
  formatCurrency(
    columns = c("2020", "2021", "2022", "2023", "2024"),
    currency = "$",
    digits = 0
  )
```
```{r, echo = FALSE}
fund_type_table %>%
  mutate(across(c("2020", "2021", "2022", "2023", "2024"), 
                ~ round(. / sum(.) * 100))) %>%
  arrange(desc(`2024`)) %>%
  datatable(
    caption = "Chicago Property Tax Levies by Fund Type — Share by Year",
    rownames = FALSE,
    options = list(
      pageLength = nrow(fund_type_table),
      dom = "t",
      columnDefs = list(
        list(className = "dt-left", targets = 0),
        list(className = "dt-right", targets = 1:5)
      )
    )
  ) %>%
  formatRound(
    columns = c("2020", "2021", "2022", "2023", "2024"),
    digits = 1
  ) %>%
  formatString(
    columns = c("2020", "2021", "2022", "2023", "2024"),
    suffix = "%"
  )
```

